  
  
```{r}
library(tidyverse)
library(ggrepel)
library(ggimage)
library(nflfastR)
library(dplyr)
options(scipen = 9999)

load(file = "../output/data.sos1.RData")

data.sos <- data.sos[,c(1:110)]
```



```{r}
# =========================
# Add ratio/efficiency features
# =========================
library(dplyr)
library(tidyr)

add_ratio_features <- function(df) {
  # Safe division: returns NA if denom <= 0 or NA
  safe_div <- function(num, den) ifelse(is.na(num) | is.na(den) | den <= 0, NA_real_, num / den)

  df %>%
    mutate(
      # --- Play-level efficiency ---
      yds_per_play          = safe_div(yards_gained, play),
      yds_per_rush          = safe_div(rushing_yards, rush_attempt),
      yds_per_pass_att      = safe_div(passing_yards, pass_attempt),
      comp_pct              = safe_div(complete_pass, pass_attempt),
      sack_rate             = safe_div(sack, qb_dropback),
      int_rate              = safe_div(interception, pass_attempt),
      pass_td_rate          = safe_div(pass_touchdown, pass_attempt),
      rush_td_rate          = safe_div(rush_touchdown, rush_attempt),

      # --- Drive-level rates ---
      scoring_drive_pct     = safe_div(drive_ended_with_score, drive_summary),
      turnover_drive_pct    = safe_div(drive_ended_with_turnover, drive_summary),
      three_and_out_rate    = safe_div(drive_three_and_out, drive_summary),
      punt_rate             = safe_div(drive_ended_with_punt, drive_summary),
      missed_fg_drive_pct   = safe_div(drive_ended_with_missedFG, drive_summary),
      
      # --- First downs & situational splits ---
      first_down_conv_rate  = safe_div(first_down, first_down_plays + second_down_plays + third_down_plays + fourth_down_plays),
      run_pass_ratio        = safe_div(rush_attempt, rush_attempt + pass_attempt),  # share of plays that are runs
      first_down_rush_share = safe_div(first_down_rush, first_down),
      first_down_pass_share = safe_div(first_down_pass, first_down),
      first_down_pen_share  = safe_div(first_down_penalty, first_down),

      # --- Passing quality / QB pressure ---
      yac_per_comp          = safe_div(yards_after_catch, complete_pass),
      air_yds_per_att       = safe_div(air_yards, pass_attempt),
      air_to_yac_ratio      = ifelse(is.na(yards_after_catch) | yards_after_catch == 0, NA_real_, air_yards / yards_after_catch),
      epa_per_dropback      = safe_div(epa, qb_dropback),
      qb_hit_rate           = safe_div(qb_hit, qb_dropback),

      # --- EPA/WPA flavor (optional, often predictive) ---
      wpa_per_play          = safe_div(wpa, play),
      qb_epa_per_dropback   = safe_div(qb_epa, qb_dropback),
      air_epa_per_att       = safe_div(air_epa, pass_attempt),
      yac_epa_per_comp      = safe_div(yac_epa, complete_pass),

      # --- Penalties & turnovers ---
      pen_yds_per_play      = safe_div(penalty_yards, play),
      fumble_rate           = safe_div(fumble, play),
      fumble_lost_rate      = safe_div(fumble_lost, fumble),
      forced_fumble_share   = safe_div(fumble_forced, fumble_forced + fumble_not_forced),

      # --- Red zone (inside the 20) heuristics ---
      # Trips = drives that reach I20. Offensive TDs = rush + pass TDs.
      rz_trips              = drive_inside20,
      rz_td                 = pass_touchdown + rush_touchdown,
      rz_td_rate            = safe_div(rz_td, rz_trips),
      rz_int_rate           = safe_div(Ints_Off_Inside20, rz_trips),

      # --- Score & line normalization ---
      points_for_per_play   = safe_div(posteam_score, play),
      points_allowed_per_play = safe_div(defteam_score, play),
      cover_margin          = posteam_score - defteam_score + spread_line,   # (>0 means covered)
      total_margin          = posteam_score + defteam_score - total_line,    # (>0 means over)

      # --- Balance ---
      rush_share            = safe_div(rush_attempt, play),
      pass_share            = safe_div(pass_attempt, play),
      rush_vs_pass_ratio    = safe_div(rush_attempt, pass_attempt),

      # --- Pressure/neg plays per play ---
      tfl_rate              = safe_div(tackled_for_loss, play),
      sack_tfl_rate         = safe_div(sack + tackled_for_loss, play)
    ) %>%
    # Optional: clean up any NaN/Inf (shouldn’t exist with safe_div, but just in case)
    mutate(across(where(is.numeric), ~ ifelse(is.nan(.x) | is.infinite(.x), NA_real_, .x)))
}
```


```{r}
# =========================
# Usage
# =========================
data.sos <- data.sos %>% add_ratio_features()

```



```{r}
library(dplyr)

# Safe division helper (same as before)
safe_div <- function(num, den) ifelse(is.na(num) | is.na(den) | den <= 0, NA_real_, num / den)

# Extend your feature engineering
data.sos <- data.sos %>%
  mutate(
    # -----------------------------
    # Situational & red-zone
    # -----------------------------
    third_down_conv_rate   = safe_div(first_down, (third_down_plays + fourth_down_plays)),
    early_vs_late_split    = safe_div(SuccOff_earlydown, SuccOff_latedown),

    # You already have rz_td_rate and rz_int_rate (per trip). Keep them,
    # and add a balance ratio TDs vs INTs (handles 0 safely).
    rz_td_to_int_ratio     = ifelse(is.na(rz_int_rate) | rz_int_rate == 0, NA_real_, rz_td_rate / rz_int_rate),

    # Points normalizations
    points_per_rz_trip     = safe_div(posteam_score, rz_trips),
    points_per_drive       = safe_div(posteam_score, drive_summary),
    points_per_100_yards   = safe_div(posteam_score, (yards_gained / 100)),

    # -----------------------------
    # Passing composition
    # -----------------------------
    air_vs_yac_dominance   = safe_div(air_yds_per_att, (air_yds_per_att + yac_per_comp)),

    # -----------------------------
    # Turnover/pressure composites
    # -----------------------------
    turnover_worthy_play_rate = int_rate + fumble_rate,
    negative_play_rate      = safe_div((sack + tackled_for_loss + interception + fumble), play),
    pressure_to_sack_conv   = safe_div(sack, (sack + qb_hit)),

    # -----------------------------
    # Betting-oriented transforms
    # -----------------------------
    # score_diff = PF - PA
    score_diff              = posteam_score - defteam_score,
    spread_adjusted_perf    = score_diff - spread_line,                      # (= cover_margin but framed explicitly)
    total_adjusted_scoring  = safe_div((posteam_score + defteam_score), total_line),

    # -----------------------------
    # Efficiency × usage interactions
    # -----------------------------
    rush_eff_x_share        = yds_per_rush * rush_share,
    pass_epa_x_share        = epa_per_dropback * pass_share,
    fd_conv_x_score_drive   = first_down_conv_rate * scoring_drive_pct,
    
    # Weighted efficiencies
    rush_eff_weighted   = yds_per_rush      * rush_share,
    pass_eff_weighted   = epa_per_dropback  * pass_share,

    # Turnovers / negative plays
    neg_play_rate       = safe_div((sack + tackled_for_loss + interception + fumble), play),
    turnover_play_rate  = int_rate + fumble_rate,
    
    #Explosive play rate
    explosive_play_rate = explosive_plays / play
  )

```


  
  
##########################################################################################
Why these help

balance_tilt / eff_weight_gap: direction and size of your offensive tilt (pass vs rush), accounting for usage and efficiency.

red_zone_leverage / finishing_vs_yards / explosive_finishing: separate yard accumulation from scoring conversion, a common blind spot.

sustain_vs_explosive / early_late_synergy: teams that can both convert 3rd downs and stay efficient late tend to outperform lines.

ball_security_risk / pressure_ composites*: capture compounding effects of pressure and turnover-worthy tendencies.

spread_signal_weighted / over_signal: marry market-relative performance with how your offense scores (sustain/explosive) to predict spread/total deviations.

logit & log transforms: make bounded rates friendlier for GLM/elastic-net, improving calibration and interpretability.

```{r}
library(dplyr)

# Helpers
safe_div <- function(a, b) ifelse(is.na(a) | is.na(b) | b == 0, NA_real_, a / b)
clamp01  <- function(p, eps = 1e-6) pmin(pmax(p, eps), 1 - eps)
logit    <- function(p) log(clamp01(p) / (1 - clamp01(p)))

data.sos <- data.sos %>%
  mutate(
    # ---------------------------
    # Balance / tilt signals
    # ---------------------------
    balance_tilt              = pass_epa_x_share - rush_eff_x_share,     # pass-lean minus rush-lean (usage-weighted)
    eff_weight_gap            = pass_eff_weighted - rush_eff_weighted,   # efficiency-weight gap

    # ---------------------------
    # Finishing & “how” a team scores
    # ---------------------------
    #red_zone_leverage         = points_per_rz_trip * rz_td_to_int_ratio, # finishing power in RZ
    finishing_vs_yards        = safe_div(points_per_drive, points_per_100_yards), # drive-finishing vs yard-churn
    explosive_finishing       = explosive_play_rate * points_per_drive,  # do explosives translate into points?

    # ---------------------------
    # Sustained vs explosive offense
    # ---------------------------
    sustain_vs_explosive      = safe_div(third_down_conv_rate, explosive_play_rate),
    early_late_synergy        = early_vs_late_split * third_down_conv_rate,

    # ---------------------------
    # Risk, pressure, & drive killers
    # ---------------------------
    ball_security_risk        = turnover_play_rate + neg_play_rate,      # broad turnover/neg-play risk
    pressure_turnover_couple  = pressure_to_sack_conv * turnover_worthy_play_rate,
    pressure_damage           = pressure_to_sack_conv * negative_play_rate,
    clutch_factor             = third_down_conv_rate * fd_conv_x_score_drive,

    # ---------------------------
    # Betting-oriented composites
    # ---------------------------
    spread_signal_weighted    = spread_adjusted_perf * fd_conv_x_score_drive,             # spread signal × sustain/score ability
    over_signal               = total_adjusted_scoring * (pass_epa_x_share + explosive_play_rate),

    # ---------------------------
    # Log / logit transforms (often helpful for linear models)
    # ---------------------------
    third_down_logit          = logit(third_down_conv_rate),
    explosive_logit           = logit(explosive_play_rate),
    p2s_logit                 = logit(pressure_to_sack_conv),
    air_vs_yac_logit          = logit(air_vs_yac_dominance),
    ln_points_per_rz_trip     = log1p(points_per_rz_trip),
    ln_points_per_drive       = log1p(points_per_drive)
  )



```


```{r}

library(dplyr)

# helper
safe_rate <- function(num, den) ifelse(is.na(num) | is.na(den) | den <= 0, NA_real_, num / den)

data.sos <- data.sos %>%
  mutate(
    # robust RZ rates (fallback to counts if rate cols are missing/NA)
    td_rate_rz  = dplyr::coalesce(rz_td_rate,  safe_rate(rz_td, rz_trips)),
    int_rate_rz = dplyr::coalesce(rz_int_rate, safe_rate(Ints_Off_Inside20, rz_trips)),

    # epsilon smoothing tied to sample size
    eps_rate = 0.5 / pmax(rz_trips, 1),

    # smoothed ratio and final leverage
    rz_td_to_int_ratio_eps = (td_rate_rz + eps_rate) / (int_rate_rz + eps_rate),
    red_zone_leverage = points_per_rz_trip * rz_td_to_int_ratio_eps
  ) 


```




```{r}
library(dplyr)

data.test <- data.sos %>% 
  filter(season != 2025) %>%
  filter(season != 2014)

na_summary_by_season <- data.test %>%
  group_by(season) %>%
  summarise(across(everything(), ~sum(is.na(.)), .names = "na_{.col}"))


#write.csv(data.test, "data.test.csv")
```














```{r}
library(tidyr)

na_long <- na_summary_by_season %>%
  pivot_longer(-season, names_to = "variable", values_to = "na_count")
```





```{r}

# columns to impute
cols_to_fix <- c("red_zone_leverage", "ln_points_per_rz_trip", "p2s_logit", "pressure_damage", "pressure_turnover_couple", "sustain_vs_explosive", "finishing_vs_yards", "pressure_to_sack_conv", "points_per_rz_trip", "rz_td_rate", "rz_int_rate","fumble_lost_rate", "forced_fumble_share", "rz_td_to_int_ratio", "td_rate_rz", "early_late_synergy", "int_rate_rz", "rz_td_to_int_ratio_eps")

# (A) Impute: replace NA with column mean (if a column is all NA, leave as NA)
reference_rows <- !is.na(data.sos$season) & data.sos$season < 2026
stopifnot(any(reference_rows))
for (col in cols_to_fix) {
  m <- mean(data.sos[[col]][reference_rows], na.rm = TRUE)
  if (is.finite(m)) {
    data.sos[[col]][is.na(data.sos[[col]])] <- m
  }
}

# (B) Check remaining NA counts (corrected)
colSums(is.na(data.sos[cols_to_fix]))


```

```{r}

data.test <- data.sos %>% 
  filter(season != 2025) %>%
  filter(season != 2014)


colSums(is.na(data.test))
```



```{r}
write.csv(data.sos, file="../output/data.sos2.csv")
 
save(data.sos, file = "../output/data.sos2.RData")
``` 
 
